Diffusion Models for Tissues and Tumor Growth

### 2024-03-16 paper.tex
Rough draft finished

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{tikz}
\title{ODE and Diffusion Models of Tumor Growth}
\author{Daniel Henricks}

@@

Researchers have been studying cancer and the patterns behind tumor growth for thousands of years. Hippocrates
was accredited with possibly being the first to describe a tumor, which were mainly described by their color
and their response to touch. However, ancient physicians did not separate benign and malignant tumors, but they
acknowledged the tumor's ability to infect nearby tissue. The application of mathematical models to tumor growth
did not start to gain traction until much later, however. In the early 20th century, one of the most famous uses of
the diffusion model was first developed: Hill released a paper discussing how the movement of lactic acid
and oyxgen throughout the body could be shaped into a diffusion problem (Hill, 1928). This paper eventually led to
Greenspan's paper which modeled the growth of tumors using diffusion, applying many of the mathematical tools employed
by Hill some years before (Greenspan, 1972). This paper will present a brief overview of the mathematics used in these
papers and will recreate some of the graphics produced as well.

\section{Hill's Paper}

@@

\end{document}

>>>>

\documentclass{article}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{tikz}
\title{Diffusion Models for Tissues and Tumor Growth}
\author{Daniel Henricks}

@@

Researchers have been studying cancer and the patterns behind tumor growth for thousands of years. Hippocrates
was accredited with possibly being the first to describe a tumor, which were mainly described by their color
and their response to touch. However, ancient physicians did not separate benign and malignant tumors, but they
acknowledged the tumor's ability to infect nearby tissue. The application of mathematical models to tumor growth
did not start to gain traction until much later, however. In the early 20th century, one of the most famous uses of
the diffusion model was first developed: Hill released a paper discussing how the movement of lactic acid
and oyxgen throughout the body could be shaped into a diffusion problem (Hill, 1928). This paper eventually led to
Greenspan's paper which modeled the growth of tumors using diffusion, applying many of the mathematical tools employed
by Hill some years before (Greenspan, 1972). This paper will present an overview of the diffusion model in mathematical
biology and will discuss the mathematics and biology behind two of the most influential diffusion papers.

\section{Hill's Paper on Diffusion}

@@

The equations (4.6) - (4.9) will be verified by checking the solutions. The equations given in Greenspan (1972) are as follows:

If $r_1(t) < r < r_3(t)$:
\begin{equation}
    \sigma = \sigma_{\infty} - \frac{a}{6k}(r_3^2 - r^2) + \frac{ar_1^3}{3k} (\frac{1}{r} - \frac{1}{r_3})
\end{equation}

\begin{equation}
    \beta = \frac{Pr_1^3}{3k'}(\frac{1}{r} - \frac{1}{r_3})
\end{equation}

From the boundaries of the step functions found above, in this region, evaluating the left-hand side of $\beta$ should
yield zero, since $r > r_1(t)$.

First checking $\beta$:
$$
    \frac{1}{r^2} \frac{\partial}{\partial r} r^2 \frac{\partial}{\partial r} (\frac{Pr_1^3}{3k'}(\frac{1}{r} - \frac{1}{r_3}))
$$

$$
    = \frac{1}{r^2} \frac{\partial}{\partial r} r^2 (-\frac{Pr_1^3}{3kr^2})
$$

$$
    = \frac{1}{r^2} (\* 0) = 0
$$

Next, checking $\sigma$:

$$
    \frac{1}{r^2} \frac{\partial}{\partial r} r^2 \frac{\partial}{\partial r} [\sigma_{\infty} - \frac{a}{6k}(r_3^2 - r^2) + \frac{ar_1^3}{3k} (\frac{1}{r} - \frac{1}{r_3})]
$$

$\sigma_{\infty}$ does not depend on the value of $r$, so it is a constant. We get

$$
    \frac{1}{r^2} \frac{\partial}{\partial r} r^2 (\frac{ar}{3k} - \frac{ar_1^3}{3kr^2})
$$

$$
    = \frac{1}{r^2} (\frac{ar^2}{k})
$$

$$
    = \frac{a}{k}
$$

So we have successfully shown that the first two equations hold according to the parameters provided via the assumptions.
Now doing the same process for equations (4.8) and (4.9):

If $ r \leq r_1(t)$:

\begin{equation}
    \sigma = \sigma_{min}
\end{equation}

\begin{equation}
    \beta = \frac{Pr_1^3}{3k'}(\frac{3}{2r_1} - \frac{r^2}{2r_1^3} - \frac{1}{r_3})
\end{equation}

The verification of $\sigma$ can be done directly. If $r < r_1(t)$, the only solution that exists must be $\sigma = \sigma_{min}$.
This is because of being inside of the necrotic core and has been shown in equation (17).

To verify $\beta$, the same method as before will be used. We start with the equation

$$
    \frac{1}{r^2} \frac{\partial}{\partial r} r^2 \frac{\partial}{\partial r} [\frac{Pr_1^3}{3k'}(\frac{3}{2r_1} - \frac{r^2}{2r_1^3} - \frac{1}{r_3})]
$$

$$
    = \frac{1}{r^2} \frac{\partial}{\partial r} r^2 (-\frac{Pr}{3k'})
$$

$$
    = \frac{1}{r^2} (-\frac{Pr^2}{k'})
$$

$$
    = -\frac{P}{k'}
$$

All of the differential equations have now been verified. We can solve for $\sigma_{min}$ and $\beta_{max}$:

First, solving for $\sigma_{min}$ requires using equation (17):

$$
    \sigma(r_1(t), t) = \sigma_{min}
$$

We now have solutions to the differential equations and directly plug in to them. Since $ r = r_1(t)$, the first equation is used.

$$
    \sigma(r_1(t), t) = \sigma_{\infty} - \frac{a}{6k}(r_3^2 - r_1^2) + \frac{ar_1^3}{3k} (\frac{1}{r_1} - \frac{1}{r_3})
$$

After simplification, this is
\begin{equation}
    \sigma_{\infty} - \sigma_{min} = \frac{a}{3k}[\frac{r_3^2-r_1^2}{2} - r_1^2 + \frac{r_1^3}{r_3}]
\end{equation}

For $\beta$, equation (16) is used:

$$
    \beta(r_2(t), t) = \beta_{max}
$$

Then the same process for finding the max value is used:

\begin{equation}
    \beta_{max} = \frac{P}{3k'}r_1^3(\frac{1}{r_2}-\frac{1}{r_3})
\end{equation}
which represents the condition for which the slowing of mitosis for cells in the middle layer occurs.

\section{Conclusion}

Hill (1928) and Greenspan (1972) both proposed diffusion models to solve their specific modeling situation.
Many modern diffusion models now follow the patterns shown in these two papers and expand on them by adding additional
terms to the diffusion PDE. For example, a recent paper by Stepien et al. (2015) included a diffusion model for in vitro
glioblastoma growth which included the following diffusion equation:

\begin{equation}
    \frac{\partial u_i(r, t)}{\partial t} = \underset{\text{Diffusion}}{\underbrace{k \nabla^2 u_i}} + \underset{\text{Logistic growth}}{\underbrace{g u_i \left(1 - \frac{u_i}{u_{\max}}\right)}} - \underset{\text{Taxis}}{\underbrace{\nu_i \nabla_r \cdot u_i}} + \underset{\text{Shed cells from core}}{\underbrace{s \delta(r - R(t))}}
\end{equation}

Many additional papers also exists which discuss how to modify Fick's second law equation in order to better model biological processes. Possible improvments to the two models above could be adding a logistic growth term into the diffusion equation; however, both authors
were limited by the lack of computing power and availability of past models. Both authors utilized similar methods for completely different
biological topics, demonstrating the versatility of the diffusion model. Overall, the diffusion model is a relatively simple and flexible model to apply to
various biological circumstances.

\begin{thebibliography}{99}

    \bibitem{citation-key4}
    Araujo, R. P., McElwain, D. L. S. (2004).
    \newblock A Historu of the Study of Solid Tumor Growth: The Contribution of Mathematical Modeling,
    \newblock \emph{Bulletin of Mathematical Biology},
    \newblock Volume~66, pp 1039--1091.

    \bibitem{citation-key2}
    Burton, A. C. (1966).
    \newblock Rate of Growth of Solid Tumors as a Problem of Diffusion,
    \newblock \emph{Growth},
    \newblock Volume~30, pp 157--176.

    \bibitem{citation-key3}
    Greenspan, H. P. (1972).
    \newblock Models for the Growth of a Solid Tumor by Diffusion,
    \newblock \emph{Stud. Appl. Math},
    \newblock Volume~52, pp 317--340.

    \bibitem{citation-key1}
    Hill, A. V. (1928).
    \newblock The Diffusion of Oxygen and Lactic Acid through Tissues,
    \newblock \emph{R. Soc Proc. B},
    \newblock Volume~104, No.~728, pp 39--96.

    \bibitem{citation-key5}
    Stepien, T. L., Rutter, E. M., Kuang, Y. (2015).
    \newblock A Data-Motivated Density-Dependent Diffusion Model of In Vitro Glioblastoma Growth,
    \newblock \emph{Stud. Appl. Math},
    \newblock Volume~12, No.~6, pp 1157--1172.

\end{thebibliography}
\end{document}
